---
title: "Course Notes 4"
author: "Brooke Anderson, Assistant Professor of Epidemiology"
job: Colorado State University
logo        : figures/CSU_ram.png
date: "January 12, 2016"
output: ioslides_presentation
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
mode        : selfcontained # {standalone, draft}
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(fig.path='figures/notes4/plot-')

library(ggplot2)
library(dplyr)
library(tidyr)
library(faraway)
data(worldcup)
data(nepali)
```

# Data wrangling

## Data wrangling

- Data cleaning
- Data wrangling
- Data munging

## Data wrangling packages

- `dplyr`: Working with a single dataset
- `tidyr`: Combining datasets

Tips on both in RStudio's [Data Wrangling Cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)

## Data wrangling

Learning these tools will take practice. For more examples, see:

- [dplyr tutorial](http://genomicsclass.github.io/book/pages/dplyr_tutorial.html)
- [RStudio intro to dplyr](https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html)
- [dplyr intro](http://www.sharpsightlabs.com/dplyr-intro-data-manipulation-with-r/)

## `dplyr`

- `select`: Select certain columns
- `filter`: Select certain rows (like `subset`)
- `mutate`: Create a new variable
- `summarize`: Summarize data (either by a variable or overall)
- `group_by`: Group data by a variable
- `arrange`: Re-order data

## `select`

Select certain columns:

```{r}
nepali[1:2, ]
select(nepali, id, sex, wt, ht, age)[1:2, ]
```

## `select`

Use `-` to exclude columns:

```{r}
nepali[1:2, ]
select(nepali, -mage, -lit, -died, -alive)[1:2, ]
```

## `filter`

Select certain rows (like `subset`):

```{r}
nepali[1:2, ]
filter(nepali, sex == 2)[1:2, ]
```

## `mutate`

Create a new variable:

```{r}
nepali[1:2, ]
mutate(nepali, ht_wt_ratio = ht / wt)[1:2, ]
```

## `summarize`

Summarize data overall:

```{r}
nepali[1:2, ]
summarize(nepali,
          wt = mean(wt, na.rm = TRUE),
          ht = mean(ht, na.rm = TRUE))
```

## `group_by` and `summarize`

Summarize data by group:

```{r}
nepali[1:2, ]
summarize(group_by(nepali, id),
          wt = mean(wt, na.rm = TRUE),
          ht = mean(ht, na.rm = TRUE))[1:2, ]
```

## `arrange`

Re-order data:

```{r}
nepali[1:2, ]
arrange(nepali, desc(wt))[1:2, ]
```

## `gather` / `spread`

You can also use two functions from the `tidyr` package to change between wide and long data: `gather()` and `spread()`. (Example from `tidyr` help files.)

Here is some wide data:

```{r, include = FALSE}
wide_stocks <- data.frame(
  time = as.Date('2009-01-01') + 0:9,
  X = rnorm(10, 0, 1),
  Y = rnorm(10, 0, 2),
  Z = rnorm(10, 0, 4)
)
```

```{r}
wide_stocks[1:3, ]
```

## `gather` / `spread`

`gather` to convert wide data to long data:

```{r}
long_stocks <- gather(wide_stocks, stock, price, -time)
long_stocks[1:5, ]
```


## `gather` / `spread`

`spread` to convert long data to wide data:

```{r}
stocks <- spread(long_stocks, stock, price)
stocks[1:5, ]
```

## Chaining / piping

To write cleaner code, `dplyr` includes the option to "pipe" or "chain" using `%>%`. This sends the output from the function before the `%>%` in as input to the function after the `%>%`.

For example:

```{r}
group_by(worldcup, Team) %>%
        summarize(max_time = max(Time)) %>%
        filter(max_time > 520)
```

# Using `dplyr` functions to tidy data

## Tidy `nepali` data

Say we wanted to tidy up the data by:

- Move variables that are constant for each child across all measurements (e.g., `mage`, `lit`, `died`, `alive`) to another dataset
- Determine each child's age at first measurement
- Limit the measurement dataset with just males
- Add new variables for (1) height-to-weight ratio and (2) months since first measurement

## Tidy `nepali` data

Move variables that are constant for each child across all measurements (e.g., `mage`, `lit`, `died`, `alive`) to another dataset:

```{r}
child_constants <- select(nepali, id, mage, lit, died, alive) %>%
        group_by(id) %>%
        summarize(mage = mean(mage), lit = mean(lit), 
                  died = mean(died), alive = mean(alive))
child_constants[1:2, ]
```

## Tidy `nepali` data

Determine each child's age at first measurement:

```{r}
first_age <- group_by(nepali, id) %>%
        summarize(first_age = min(age))
first_age[1:2, ]
```

## Tidy `nepali` data

- Limit the measurement dataset with just males
- Add new variables for (1) height-to-weight ratio and (2) months since first measurement

```{r}
child_measures <- select(nepali, -mage, -lit, -died, -alive) %>%
        filter(sex == 1) %>%
        left_join(first_age, by = "id") %>%
        mutate(ht_wt_ratio = ht / wt, 
               months = age - first_age)

child_measures[1:2, ]
```

# `*_join` functions

## `*_join` functions

The `dplyr` package has a family of different functions to join two dataframes together. All combine two dataframes, which I'll call `x` and `y` here. The functions include:

- `inner_join(x, y)`: Keep only rows where there are observations in both `x` and `y`.
- `left_join(x, y)`: Keep all rows from `x`, whether they have a match in `y` or not. 
- `right_join(x, y)`: Keep all rows from `y`, whether they have a match in `x` or not. 
- `full_join(x, y)`: Keep all rows from both `x` and `y`, whether they have a match in the other dataset or not.

## `*_join` functions

```{r}
(x <- data.frame(course = c("x", "y", "z"),
                grade = c(90, 82, 78)))
(y <- data.frame(course = c("w", "x", "y"),
                day = c("Tues", "Mon / Fri", "Tue")))
```

## `*_join` functions

With `inner_join`, you'll lose data on `z` (only in the first dataset) and `w` (only in the second dataset). You'll only get the observations that show up in both datasets:

```{r warning = FALSE}
inner_join(x, y)
```

## `*_join` functions

With `left_join`, you'll keep everything in `x` (the "left" dataset), but not keep things in `y` that don't match something in `x`. That means that, here, you'll lose `w`:

```{r, warning = FALSE}
left_join(x, y)
```

## `*_join` functions

`right_join` is the opposite:

```{r, warning = FALSE}
right_join(x, y)
```

## `*_join` functions

`full_join` keeps everything from both datasets:

```{r warning = FALSE}
full_join(x, y)
```
