---
title: "Plotting"
author: "Brooke Anderson, Assistant Professor of Epidemiology"
job: Colorado State University
logo        : figures/CSU_ram.png
date: "January 12, 2016"
output: ioslides_presentation
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
mode        : selfcontained # {standalone, draft}
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
knitr::opts_chunk$set(fig.path='figures/notes2/plot-')

library(ggplot2)
library(faraway)
library(dlnm)
library(gridExtra)
library(ggthemes)
library(dplyr)

data(worldcup)
data(chicagoNMMAPS)
chic <- chicagoNMMAPS
```

# Guidelines for good plots

## High data density

* Guideline 1: **Aim for high data density.**

**Data to ink ratio**: 

Ratio of "ink" providing information to all ink

## High data density

* Guideline 1: **Aim for high data density.**

```{r echo = FALSE, fig.height = 3, fig.width = 8}
a <- ggplot(worldcup, aes(Position)) + 
        geom_bar() + coord_flip() + 
        ylab("Number of players") + 
        ggtitle("1. Lower data density")

ex <- group_by(worldcup, Position) %>%
        summarise(n = n())
b <- ggplot(ex, aes(x = n, y = Position)) + 
        geom_point() +  
        xlab("Number of players") + 
        theme_few() + 
        xlim(0, 250) + 
        ggtitle("2. Higher data density")
grid.arrange(a, b, ncol = 2)
```

## High data density

* Guideline 1: **Aim for high data density.**

```{r echo = FALSE, fig.height = 3, fig.width = 8}
chic_july <- subset(chic, month == 7 & year == 1995)
a <- ggplot(chic_july, aes(x = date, y = death)) + 
        geom_area(fill = "black") + 
        xlab("Day in July 1995") + 
        ylab("All-cause deaths") + 
        ylim(0, 450) + 
        theme_excel() + 
        ggtitle("1. Lower data density")

b <- ggplot(chic_july, aes(x = as.POSIXlt(date)$mday,
                           y = death)) + 
        geom_line() + 
        xlab("Day in July 1995") + 
        ylab("All-cause deaths") + 
        ylim(0, 450) + 
        theme_tufte() + 
        ggtitle("2. Higher data density")

grid.arrange(a, b, ncol = 2)
```

## Meaningful labels

* Guideline 2: **Use clear, meaningful labels.**

```{r echo = FALSE, fig.height = 3, fig.width = 8}
ex <- worldcup
ex$Position <- factor(ex$Position, 
                      levels = c("Defender",
                                 "Forward",
                                 "Goalkeeper",
                                 "Midfielder"),
                      labels = c("DF", "FW",
                                 "GK", "MF"))
a <- ggplot(ex, aes(Position)) + 
        geom_bar() + 
        ylab("Pls") + 
        xlab("Pos") +
        ggtitle("1. Murkier labels") + 
        theme(axis.text.x = 
                      element_text(angle = 90,
                                   vjust = 0.5,
                                   hjust=1))

b <- ggplot(worldcup, aes(Position)) + 
        geom_bar(fill = "lightgray") + coord_flip() + 
        ylab("Number of players") + 
        theme_tufte() + 
        ggtitle("2. Clearer labels")

grid.arrange(a, b, ncol = 2)
```

## References

* Guideline 3: **Provide useful references.**

```{r echo = FALSE, fig.height = 3, fig.width = 8}
chic_july <- subset(chic, month == 7 & year == 1995)
chic_july_ref <- filter(chic, month == 7 & 
                        year %in% c(1990:1994,
                                    1996:2000)) %>%
        summarise(mean = mean(death),
                  min = min(death),
                  max = max(death))
ref_points <- data.frame(date = c(-2, 33, 33, -2),
        death = c(rep(chic_july_ref$max, 2),
                  rep(chic_july_ref$min, 2)))
        
a <- ggplot(chic_july, aes(x = as.POSIXlt(date)$mday,
                           y = death)) + 
        geom_line() + 
        xlab("Day in July 1995") + 
        ylab("All-cause deaths") + 
        ylim(0, 450) + 
        theme_tufte() + 
        ggtitle("1. No reference")

b <- ggplot(chic_july, aes(x = as.POSIXlt(date)$mday,
                           y = death)) + 
        xlab("Day in July 1995") + 
        ylab("All-cause deaths") + 
        ylim(0, 450) + 
        theme_tufte() + 
        geom_polygon(aes(x = date, y = death), 
                     data = ref_points,
                     color = "lightgray", 
                     alpha = 0.1) + 
        geom_line() + 
        ggtitle("2. Reference")

grid.arrange(a, b, ncol = 2)
```

## References

* Guideline 3: **Provide useful references.**

```{r echo = FALSE, message = FALSE, fig.width = 8, fig.height = 4}
ex <- subset(worldcup, Position == "Forward")
a <- ggplot(ex, aes(x = Passes, y = Shots)) + 
        geom_point(size = 1.5) + 
        theme_few()  + 
        ggtitle("1. No reference")

b <- ggplot(ex, aes(x = Passes, y = Shots)) + 
        geom_point(size = 1.5) + 
        theme_few()  + 
        geom_smooth() + 
        ggtitle("2. Reference")

grid.arrange(a, b, ncol = 2)
```

## Interesting aspects

* Guideline 3: **Highlight interesting aspects.**

```{r echo = FALSE, fig.height = 3, fig.width = 8}
chic_july <- subset(chic, month == 7 & year == 1995)
chic_july_ref <- filter(chic, month == 7 & 
                        year %in% c(1990:1994,
                                    1996:2000)) %>%
        summarise(mean = mean(death),
                  min = min(death),
                  max = max(death))
ref_points <- data.frame(date = c(-2, 33, 33, -2),
        death = c(rep(chic_july_ref$max, 2),
                  rep(chic_july_ref$min, 2)))
hw <- data.frame(date = c(12, 16, 16, 12),
                 death = c(425, 425, 0, 0))
        
a <- ggplot(chic_july, aes(x = as.POSIXlt(date)$mday,
                           y = death)) + 
        xlab("Day in July 1995") + 
        ylab("All-cause deaths") + 
        ylim(0, 450) + 
        theme_tufte() + 
        geom_polygon(aes(x = date, y = death), 
                     data = ref_points,
                     color = "lightgray", 
                     alpha = 0.1) + 
        geom_line() + 
        ggtitle("1. No highlighting")

b <- ggplot(chic_july, aes(x = as.POSIXlt(date)$mday,
                           y = death)) + 
        xlab("Day in July 1995") + 
        ylab("All-cause deaths") + 
        ylim(0, 450) + 
        theme_tufte() + 
        geom_polygon(aes(x = date, y = death), 
                     data = ref_points,
                     color = "lightgray", 
                     alpha = 0.1) + 
        geom_line(aes(x = date, y = death),
                  data = hw[1:2, ],
                     color = "red",
                  size = 2) +
        geom_line() + 
        ggtitle("2. With highlighting")

grid.arrange(a, b, ncol = 2)
```

## Interesting aspects

* Guideline 3: **Highlight interesting aspects.**

```{r echo = FALSE, message = FALSE, fig.width = 8, fig.height = 4}
ex <- subset(worldcup, Position == "Forward")
a <- ggplot(ex, aes(x = Passes, y = Shots)) + 
        geom_point(size = 1.5, alpha = 0.5) + 
        theme_few()  + 
        ggtitle("1. No highlighting")

most_shots <- ex[which.max(ex$Shots), ]
most_passes <- ex[which.max(ex$Passes), ]
b <- ggplot(ex, aes(x = Passes, y = Shots)) + 
        geom_point(size = 1.5, alpha = 0.5) + 
        theme_few()  + 
        ggtitle("2. Highlighting") + 
        geom_text(data = most_shots,
           label = paste(rownames(most_shots), ",",
                                most_shots$Team, " "),
                  colour = "blue", size = 3,
                  hjust = 1, vjust = 0.4) + 
        geom_text(data = most_passes,
           label = paste(rownames(most_passes), ",",
                                most_passes$Team, " "),
                  colour = "blue", size = 3,
                  hjust = 1, vjust = 0.4) 

grid.arrange(a, b, ncol = 2)
```

## Order

* Guideline 4: **Make order meaningful.**

```{r echo = FALSE, fig.width = 8, fig.height = 5}
ex <- group_by(worldcup, Team) %>%
        summarise(mean_time = mean(Time))

a <- ggplot(ex, aes(x = mean_time, y = Team)) + 
        geom_point() + 
        theme_few() + 
        xlab("Mean time per player (minutes)") + 
        ggtitle("1. Alphabetical order")

ex2 <- arrange(ex, mean_time) %>%
        mutate(Team = factor(Team, levels = Team))
b <- ggplot(ex2, aes(x = mean_time, y = Team)) + 
        geom_point() + 
        theme_few() + 
        xlab("Mean time per player (minutes)") + 
        ggtitle("2. Meaningful order")

grid.arrange(a, b, ncol = 2)
```

## Small multiples

* Guideline 5: **When possible, use small multiples.**

- Many small plots showing the same thing for different facets
- Same x- and y-axes
- Limit axis annotation

------

```{r echo = FALSE, message = FALSE, fig.height = 6, fig.width = 8}
ex <- subset(worldcup, Position %in% c("Forward",
                                       "Midfielder"))
ex2 <- group_by(ex, Team) %>%
        summarise(mean = mean(Shots)) %>%
        arrange(desc(mean))
ex$Team <- factor(ex$Team, 
                      levels = ex2$Team)
a <- ggplot(ex, aes(x = Time, y = Shots)) + 
        geom_point() + 
        theme_few() + 
        facet_wrap(~ Team, ncol = 8) + 
        geom_smooth(method = "lm", se = FALSE)
a
```

## Learn more

To find out more about how to make effective graphics, check out:

- Edward Tufte
- Howard Wainer
- Stephen Few
- Nathan Yau

# ggplot2

## `ggplot` conventions

Basic idea:

You'll create an object of the `ggplot` class. Then, you'll "add on" to customize it, using `+`. 

*Note*: To avoid errors, end lines with `+`, don't start lines with it. 

## `ggplot` conventions

Code conventions are different for `ggplot` than for base R graphics. 

First, you need to initialize a ggplot object using the `ggplot()` function. You can use this to specify the dataframe and map variables in your data to `x` and `y` components of the graph. 

```{r}
first_graph <- ggplot(chic_july,
                      aes(x = date, y = death))
```

This is not enough to plot something-- you will need to add one or more `geom` to plot.

## `ggplot` conventions

To add a scatterplot, you can add a `geom_point()`:

```{r, fig.width = 4, fig.height = 3}
first_graph + geom_point()
```

## `ggplot` conventions

Note that there are a few ways you could have done this:

Without saving the plot as an object:
```{r eval = FALSE}
ggplot(chic_july, aes(x = date, y = death)) + 
        geom_point()
```

Saving as an object, then save on the added geom, then print:
```{r eval = FALSE}
first_graph <- ggplot(chic_july,
                      aes(x = date, y = death))
first_graph <- first_graph + geom_point()
first_graph
```

## `ggplot` conventions

```{r, fig.width = 4, fig.height = 3, eval = FALSE}
first_graph <- ggplot(chic_july,
                      aes(x = date, y = death))
first_graph <- first_graph + geom_point()

first_graph
first_graph + geom_line() +
        xlab("Date in 1995") + 
        ylim(0, 500)
```

-----------------

```{r echo = FALSE}
first_graph <- ggplot(chic_july,
                      aes(x = date, y = death))
first_graph <- first_graph + geom_point()
```

```{r, echo = FALSE, fig.width = 4, fig.height = 3}
first_graph
first_graph + geom_line() +
        xlab("Date in 1995") + 
        ylim(0, 500)
```

# Using ggplot

## Initializing the object

Use the following conventions to initialize a ggplot object:

```{r, eval = FALSE}
## Pseudo-code, will not run
[object name] <- ggplot([dataframe],
                        aes(x = [x variable],
                            y = [y variable]))
```

## Adding geoms

Now you can add one or more `geom` to the object to create a plot. Here are some common ones:

- `geom_point()`
- `geom_line()`, `geom_segment()`
- `geom_path()`, `geom_polygon()`
- `geom_histogram()`
- `geom_abline()`, `geom_hline()`, `geom_vline()`
- `geom_boxplot()`
- `geom_smooth()`, `geom_text()`

## Adding geoms

For example:

```{r, fig.width = 3.5, fig.height = 3}
ggplot(chic_july, aes(x = temp, y = death)) + 
        geom_point()
```

## Adding geoms

For example:

```{r, fig.width = 8, fig.height = 3, message = FALSE}
ggplot(worldcup, aes(x = Time)) + 
        geom_histogram(binwidth = 10) + 
        geom_vline(xintercept = 90 * 0:6,
                   color = "blue", alpha = 0.5)
```

## Adding geoms

Common options for geoms include:

- `color` / `colour`
- `size`
- `alpha` (transparency)

## Adding geoms

Geom options can either be **constant** (e.g., all points are red), or can be **mapped** to a certain characteristic of the data:

```{r fig.width = 3, fig.height = 3}
a <- ggplot(chic_july, aes(x = date, y = death))  +
        geom_point(color = "red")

chic_july$high_temp <- chic_july$temp > 28
b <- ggplot(chic_july, aes(x = date, y = death,
                      color = high_temp))  +
        geom_point()
```

----

```{r fig.height = 3, fig.width = 8}
grid.arrange(a, b, ncol = 2)
```

## Data density

Using themes: You can change the "background" elements to a plot by changing the theme. 

The `ggthemes` packages has some excellent ones to help reduce "non-data ink".

## Data density

For example:

```{r}
chic_plot <- ggplot(chic_july,
                    aes(x = date, y = death))  +
        geom_point(color = "red")
```

## Themes

Stephen Few theme: 

```{r, fig.height = 3, fig.width = 8}
a <- chic_plot
b <- chic_plot + theme_few()
grid.arrange(a, b, ncol = 2)
```

## Themes

Edward Tufte theme: 

```{r, fig.height = 3, fig.width = 8}
a <- chic_plot
b <- chic_plot + theme_tufte()
grid.arrange(a, b, ncol = 2)
```

## Themes

Black-and-white theme: 

```{r, fig.height = 3, fig.width = 8}
a <- chic_plot
b <- chic_plot + theme_bw()
grid.arrange(a, b, ncol = 2)
```

## Themes

Excel theme: 

```{r, fig.height = 3, fig.width = 8}
a <- chic_plot
b <- chic_plot + theme_excel()
grid.arrange(a, b, ncol = 2)
```

## Highlighting

One helpful way to annotate is with text, using `geom_text()`:

```{r}
hottest_day <- chic_july[which.max(chic_july$temp),]
hottest_day
```

## Highlighting

```{r fig.height = 3, fig.width = 4}
chic_plot + geom_text(data = hottest_day, 
                      label = "Max",
                      size = 3, hjust = 0.5, vjust = 0)
```

## Highlighting

You can also use lines to highlight:

```{r}
chic_july <- subset(chic, month == 7 & year == 1995)
hw <- data.frame(date = c(as.Date("1995-07-12"), 
                          as.Date("1995-07-16")),
                 death = c(425, 425))
        
b <- chic_plot + 
        geom_line(aes(x = date, y = death),
                  data = hw[1:2, ],
                  size = 2)
```

-----

```{r fig.height = 3, fig.width = 4}
b
```

## Small multiples

You can use the facet functions to create small multiples. This separates the graph into several small graphs, one for each level of a factor.

- `facet_grid()`
- `facet_wrap()`

## Small multiples

```{r warning = FALSE, fig.width = 8, fig.height = 3}
ggplot(nepali, aes(ht, wt)) + 
        geom_point() + 
        facet_grid(. ~ sex)
```

## Small multiples

Conventions for `facet_grid`:

```{r eval = FALSE}
## Note: This is pseudocode and won't run.
facet_grid([factor for rows] ~ [factor for columns])
```

If you need a different "shape" for faceting, try `facet_wrap()` (in `gridExtra` package):

```{r eval = FALSE}
## Note: This is pseudocode and won't run.
facet_wrap(~ [factor for faceting], ncol = [# of columns])
```

## Small multiples

Re-naming and re-ordering factors:

Often, when you do faceting, you'll want to re-name your factors levels or re-order them.

For this, you'll need to use the `factor()` function on the original vector:

```{r}
nepali$sex <- factor(nepali$sex,
                     levels = c(1, 2), 
                     labels = c("Male", "Female"))
```

## Small multiples

```{r warning = FALSE, fig.width = 8, fig.height = 3}
ggplot(nepali, aes(ht, wt)) + 
        geom_point() + 
        facet_grid(. ~ sex)
```

## Small multiples

You can do the same idea to re-order a factor:

```{r}
nepali$sex <- factor(nepali$sex,
                     levels = c("Female", "Male"))
```

## Small multiples

```{r warning = FALSE, fig.width = 8, fig.height = 3}
ggplot(nepali, aes(ht, wt)) + 
        geom_point() + 
        facet_grid(. ~ sex)
```

## Annotation

You can use the following functions for main and axis titles:

- `ggtitle()`
- `xlab()`
- `ylab()`

You can also use the "convenience" functions for limits for x- and y-axes:

- `xlim()`
- `ylim()`

## Excellent references

- R Graphics Cookbook
- Google images

# Odds and ends

## R cheat sheets

- [General (CRAN)](https://cran.r-project.org/doc/contrib/Baggott-refcard-v2.pdf)
- [ggplot](https://www.rstudio.com/wp-content/uploads/2015/08/ggplot2-cheatsheet.pdf)
- [R Markdown (next week)](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)
- [Data wrangling (two weeks from now)](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
- [Data management](http://www.ualberta.ca/~ahamann/teaching/renr690/R_Cheat_Data.pdf)
- [quandl](https://s3.amazonaws.com/quandl-static-content/Documents/Quandl+-+R+Cheat+Sheet.pdf)

## Picking packages

How can I find a reliable package?

- Google "r [what I'm trying to do] tutorial pdf". Look for tutorials / lecture notes from courses at reputable universities
- Find packages with an accompanying article in the *Journal of Statistical Software* or *R Journal* (or another peer-reviewed publication)
- Use packages recommended by statistics textbooks from our library
- Look for packages referenced in peer-reviewed research in my field
- For a package for graphics, if it works, it works. For methods, be a bit more cautious.
