---
title: "Preliminaries in R"
author: "Brooke Anderson, Assistant Professor of Epidemiology"
job: Colorado State University
logo        : figures/CSU_ram.png
date: "January 12, 2016"
output: ioslides_presentation
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
mode        : selfcontained # {standalone, draft}
---

# R object types

## R's most basic object types

The two most basic types of objects for data in R are **vectors** (1D) and **dataframes** (2D). 

## Vectors

- A vector is a string of values. 
- All values must be of the same class (i.e., all numbers, all characters, all dates)
- You can use `c()` to join values together to create a vector
- The *length* of the vector is how many values it has in it

For example:
```{r}
fibonacci <- c(1, 1, 2, 3, 5)
fibonacci
length(fibonacci)
```

## Vectors

An example using characters instead of numbers:

```{r}
one_to_five <- c("one", "two", "three", "four", "five")
one_to_five
```

If you mix classes, it will default to most generic:

```{r}
mixed_classes <- c(1, 3, "five")
mixed_classes
```

## Vectors

You can pull out certain values by using indexing (`[...]`) to identify the locations you want to get

```{r}
fibonacci[2] # Get the second value
fibonacci[c(1, 5)] # Get first and fifth values
fibonacci[1:3] # Get the first three values
```

## Dataframes

A dataframe is one or more vectors of the same length stuck together side-by-side. It is the closest R has to what you'd get with an Excel spreadsheet. 

You can create dataframes using the `data.frame()` function. However, most often you will create a dataframe by reading in data from a file using something like `read.csv()`. 

## Dataframes

For example, to create a dataframe from vectors you already have saved as R objects:

```{r}
fibonacci_seq <- data.frame(num_in_seq = one_to_five,
                            fibonacci_num = fibonacci)
fibonacci_seq
```

## Dataframes

The format for using `data.frame()` is:

```{r eval = FALSE}
## Note: Generic code-- this will not work
[name of object] <- data.frame([1st column name] = [1st column content],
                               [2nd column name] = [2nd column content])
```

## Dataframes

You can use indexing (`[..., ...]`) for dataframes, too, but now they'll have two dimensions (rows, then columns). Put the rows you want before the comma, the columns after. If you want all of something, leave the designated spot blank. For example:

```{r}
fibonacci_seq[1:2, 2] # First two rows, second column
fibonacci_seq[5, ] # Last row, all columns
```

## Dataframes

Usually, instead of creating a dataframe from vectors, you'll read one in from data on an outside file. For example, to read in a dataset from a CSV file called `daily_show_guests.csv`:

```{r}
daily_show <- read.csv("data/daily_show_guests.csv",
                       header = TRUE,
                       skip = 4)
daily_show[1:2, ]
```

## Dataframes

You can use the functions `dim()`, `nrow()`, and `ncol` to figure out the dimensions (number of rows and columns) of a dataframe:

```{r}
dim(daily_show)
nrow(daily_show)
ncol(daily_show)
```

## Functions

In general, functions in R take the following structure:

```{r, eval = FALSE}
function.name(required information, options)  ## Generic code
```

The result of the function will be output to your R session, unless you choose to save the output in an object:

```{r, eval = FALSE}
new.object <- function.name(required information, options)  ## Generic code
```

## Functions

Examples of this structure:

```{r, eval = FALSE}
head(daily_show)
head(daily_show, n = 3)
daily_show <- read.csv("daily_show_guests.csv",
                    skip = 4,
                    header = TRUE)
```

Find out more about a function by using `?` (e.g., ,`?head`, `?read.csv`). This will take you to the help page for the function, where you can find out all the possible arguments for the function, required and optional.
