---
title: "Quiz 6"
author: ' '
date: ' '
output: pdf_document
---

```{r echo = FALSE, message = FALSE}
library(knitr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggthemes)
library(gridExtra)
knitr::opts_chunk$set(fig.path='figures/quiz6/plot-')
```

##### Name: 

#### `titanic` data

For all of the questions today, you will be using data for passengers from the Titanic. Here is the start of this dataset, called `titanic`:

```{r}
titanic <- read.csv("../data/titanic.csv", as.is = TRUE) %>%
        select(Survived, Pclass, Age, Name, Sex, Fare) %>% 
        mutate(Pclass = as.factor(Pclass))
head(titanic)
```


#### 1.-12. 
Match the number at the end of each line to the letter of the output you would have if you ran all code up to that point. For all dataframes, only the first few lines are shown. Put the letter answer next to the number given in the code.

```{r}
titanic$agecat[titanic$Age < 15] <- "Under 15"          # 1. 
titanic$agecat[titanic$Age >= 15 & 
                     titanic$Age <= 50] <- "15--50"     # 2. 
titanic$agecat[titanic$Age > 50] <- "Over 50"
titanic$agecat <- factor(titanic$agecat,
                         levels = c("Under 15",
                                    "15--50",
                                    "Over 50"))

toplot_4 <- filter(titanic, !is.na(Age)) %>%            # 3. 
        select(Survived, Pclass, agecat, Sex) %>%       # 4. 
        group_by(Pclass, agecat, Sex) %>%
        summarize(N = n(),
                  survivors = sum(Survived == 1),
                  perc_survived = 100 * survivors / N) %>% #5. 
        ungroup() %>%
        mutate(Pclass = factor(Pclass, levels = c(1, 2, 3),
                               labels = c("First class", "Second class",
                                          "Third class"))) # 6.
plot_4 <- ggplot(toplot_4, aes(x = agecat, y = perc_survived, fill = Pclass)) + 
        geom_bar(stat = "identity", position = "dodge") +  # 7.
        coord_flip() +                                     # 8.
        ylab("Percent survivors") + xlab("Age category") + # 9.
        theme_few() +                                      # 10.
        scale_fill_discrete(name = "Ticket class") +       # 11.
        facet_grid(. ~ Sex)                                # 12. 
```

a. 

```{r echo = FALSE, fig.height = 2, fig.width = 6}
plot_4
```

b. 

```{r echo = FALSE}
titanic <- read.csv("../data/titanic.csv", as.is = TRUE) %>%
        select(Survived, Pclass, Age, Name, Sex, Fare) %>% 
        mutate(Pclass = as.factor(Pclass))
titanic$agecat[titanic$Age < 15] <- "Under 15" 
head(titanic)
```

c. 

```{r echo = FALSE}
titanic <- read.csv("../data/titanic.csv", as.is = TRUE) %>%
        select(Survived, Pclass, Age, Name, Sex, Fare) %>% 
        mutate(Pclass = as.factor(Pclass))

titanic$agecat[titanic$Age < 15] <- "Under 15"          # 1. 
titanic$agecat[titanic$Age >= 15 & 
                     titanic$Age <= 50] <- "15--50"     # 2. 
titanic$agecat[titanic$Age > 50] <- "Over 50"
titanic$agecat <- factor(titanic$agecat,
                         levels = c("Under 15",
                                    "15--50",
                                    "Over 50"))

toplot_4 <- filter(titanic, !is.na(Age)) %>%            # 3. 
        select(Survived, Pclass, agecat, Sex) %>%       # 4. 
        group_by(Pclass, agecat, Sex) %>%
        summarize(N = n(),
                  survivors = sum(Survived == 1),
                  perc_survived = 100 * survivors / N)
head(toplot_4)
```

d. 

```{r echo = FALSE}
titanic <- read.csv("../data/titanic.csv", as.is = TRUE) %>%
        select(Survived, Pclass, Age, Name, Sex, Fare) %>% 
        mutate(Pclass = as.factor(Pclass))
titanic$agecat[titanic$Age < 15] <- "Under 15"          # 1. 
titanic$agecat[titanic$Age >= 15 & 
                     titanic$Age <= 50] <- "15--50"     # 2. 
titanic$agecat[titanic$Age > 50] <- "Over 50"
titanic$agecat <- factor(titanic$agecat,
                         levels = c("Under 15",
                                    "15--50",
                                    "Over 50"))

toplot_4 <- filter(titanic, !is.na(Age))
head(toplot_4)
```

e. 

```{r echo = FALSE}
titanic <- read.csv("../data/titanic.csv", as.is = TRUE) %>%
        select(Survived, Pclass, Age, Name, Sex, Fare) %>% 
        mutate(Pclass = as.factor(Pclass))
titanic$agecat[titanic$Age < 15] <- "Under 15"          # 1. 
titanic$agecat[titanic$Age >= 15 & 
                     titanic$Age <= 50] <- "15--50"     # 2. 
titanic$agecat[titanic$Age > 50] <- "Over 50"
titanic$agecat <- factor(titanic$agecat,
                         levels = c("Under 15",
                                    "15--50",
                                    "Over 50"))

toplot_4 <- filter(titanic, !is.na(Age)) %>%            # 3. 
        select(Survived, Pclass, agecat, Sex)
head(toplot_4)
```

f. 

```{r echo = FALSE, fig.width = 4, fig.height = 2}
titanic <- read.csv("../data/titanic.csv", as.is = TRUE) %>%
        select(Survived, Pclass, Age, Name, Sex, Fare) %>% 
        mutate(Pclass = as.factor(Pclass))
titanic$agecat[titanic$Age < 15] <- "Under 15"          # 1. 
titanic$agecat[titanic$Age >= 15 & 
                     titanic$Age <= 50] <- "15--50"     # 2. 
titanic$agecat[titanic$Age > 50] <- "Over 50"
titanic$agecat <- factor(titanic$agecat,
                         levels = c("Under 15",
                                    "15--50",
                                    "Over 50"))

toplot_4 <- filter(titanic, !is.na(Age)) %>%            # 3. 
        select(Survived, Pclass, agecat, Sex) %>%       # 4. 
        group_by(Pclass, agecat, Sex) %>%
        summarize(N = n(),
                  survivors = sum(Survived == 1),
                  perc_survived = 100 * survivors / N) %>% #5. 
        ungroup() %>%
        mutate(Pclass = factor(Pclass, levels = c(1, 2, 3),
                               labels = c("First class", "Second class",
                                          "Third class"))) # 6.
plot_4 <- ggplot(toplot_4, aes(x = agecat, y = perc_survived, fill = Pclass)) + 
        geom_bar(stat = "identity", position = "dodge") +  # 7.
        coord_flip() +                                     # 8.
        ylab("Percent survivors") + xlab("Age category") + # 9.
        theme_few()
plot_4
```

g. 

```{r echo = FALSE}
titanic <- read.csv("../data/titanic.csv", as.is = TRUE) %>%
        select(Survived, Pclass, Age, Name, Sex, Fare) %>% 
        mutate(Pclass = as.factor(Pclass))
titanic$agecat[titanic$Age < 15] <- "Under 15"          # 1. 
titanic$agecat[titanic$Age >= 15 & 
                     titanic$Age <= 50] <- "15--50"
head(titanic)
```

h. 

```{r echo = FALSE, fig.width = 5, fig.height = 2}
titanic <- read.csv("../data/titanic.csv", as.is = TRUE) %>%
        select(Survived, Pclass, Age, Name, Sex, Fare) %>% 
        mutate(Pclass = as.factor(Pclass))
titanic$agecat[titanic$Age < 15] <- "Under 15"          # 1. 
titanic$agecat[titanic$Age >= 15 & 
                     titanic$Age <= 50] <- "15--50"     # 2. 
titanic$agecat[titanic$Age > 50] <- "Over 50"
titanic$agecat <- factor(titanic$agecat,
                         levels = c("Under 15",
                                    "15--50",
                                    "Over 50"))

toplot_4 <- filter(titanic, !is.na(Age)) %>%            # 3. 
        select(Survived, Pclass, agecat, Sex) %>%       # 4. 
        group_by(Pclass, agecat, Sex) %>%
        summarize(N = n(),
                  survivors = sum(Survived == 1),
                  perc_survived = 100 * survivors / N) %>% #5. 
        ungroup() %>%
        mutate(Pclass = factor(Pclass, levels = c(1, 2, 3),
                               labels = c("First class", "Second class",
                                          "Third class"))) # 6.
plot_4 <- ggplot(toplot_4, aes(x = agecat, y = perc_survived, fill = Pclass)) + 
        geom_bar(stat = "identity", position = "dodge")
plot_4
```

i. 

```{r echo = FALSE, fig.width = 4, fig.height = 2}
titanic <- read.csv("../data/titanic.csv", as.is = TRUE) %>%
        select(Survived, Pclass, Age, Name, Sex, Fare) %>% 
        mutate(Pclass = as.factor(Pclass))
titanic$agecat[titanic$Age < 15] <- "Under 15"          # 1. 
titanic$agecat[titanic$Age >= 15 & 
                     titanic$Age <= 50] <- "15--50"     # 2. 
titanic$agecat[titanic$Age > 50] <- "Over 50"
titanic$agecat <- factor(titanic$agecat,
                         levels = c("Under 15",
                                    "15--50",
                                    "Over 50"))

toplot_4 <- filter(titanic, !is.na(Age)) %>%            # 3. 
        select(Survived, Pclass, agecat, Sex) %>%       # 4. 
        group_by(Pclass, agecat, Sex) %>%
        summarize(N = n(),
                  survivors = sum(Survived == 1),
                  perc_survived = 100 * survivors / N) %>% #5. 
        ungroup() %>%
        mutate(Pclass = factor(Pclass, levels = c(1, 2, 3),
                               labels = c("First class", "Second class",
                                          "Third class"))) # 6.
plot_4 <- ggplot(toplot_4, aes(x = agecat, y = perc_survived, fill = Pclass)) + 
        geom_bar(stat = "identity", position = "dodge") +  # 7.
        coord_flip() +                                     # 8.
        ylab("Percent survivors") + xlab("Age category") + # 9.
        theme_few() +                                      # 10.
        scale_fill_discrete(name = "Ticket class")
plot_4
```

j. 

```{r echo = FALSE, fig.width = 4, fig.height = 2}
titanic <- read.csv("../data/titanic.csv", as.is = TRUE) %>%
        select(Survived, Pclass, Age, Name, Sex, Fare) %>% 
        mutate(Pclass = as.factor(Pclass))
titanic$agecat[titanic$Age < 15] <- "Under 15"          # 1. 
titanic$agecat[titanic$Age >= 15 & 
                     titanic$Age <= 50] <- "15--50"     # 2. 
titanic$agecat[titanic$Age > 50] <- "Over 50"
titanic$agecat <- factor(titanic$agecat,
                         levels = c("Under 15",
                                    "15--50",
                                    "Over 50"))

toplot_4 <- filter(titanic, !is.na(Age)) %>%            # 3. 
        select(Survived, Pclass, agecat, Sex) %>%       # 4. 
        group_by(Pclass, agecat, Sex) %>%
        summarize(N = n(),
                  survivors = sum(Survived == 1),
                  perc_survived = 100 * survivors / N) %>% #5. 
        ungroup() %>%
        mutate(Pclass = factor(Pclass, levels = c(1, 2, 3),
                               labels = c("First class", "Second class",
                                          "Third class"))) # 6.
plot_4 <- ggplot(toplot_4, aes(x = agecat, y = perc_survived, fill = Pclass)) + 
        geom_bar(stat = "identity", position = "dodge") +  # 7.
        coord_flip() 
plot_4
```

k. 

```{r echo = FALSE}
titanic <- read.csv("../data/titanic.csv", as.is = TRUE) %>%
        select(Survived, Pclass, Age, Name, Sex, Fare) %>% 
        mutate(Pclass = as.factor(Pclass))
titanic$agecat[titanic$Age < 15] <- "Under 15"          # 1. 
titanic$agecat[titanic$Age >= 15 & 
                     titanic$Age <= 50] <- "15--50"     # 2. 
titanic$agecat[titanic$Age > 50] <- "Over 50"
titanic$agecat <- factor(titanic$agecat,
                         levels = c("Under 15",
                                    "15--50",
                                    "Over 50"))

toplot_4 <- filter(titanic, !is.na(Age)) %>%            # 3. 
        select(Survived, Pclass, agecat, Sex) %>%       # 4. 
        group_by(Pclass, agecat, Sex) %>%
        summarize(N = n(),
                  survivors = sum(Survived == 1),
                  perc_survived = 100 * survivors / N) %>% #5. 
        ungroup() %>%
        mutate(Pclass = factor(Pclass, levels = c(1, 2, 3),
                               labels = c("First class", "Second class",
                                          "Third class")))
head(toplot_4)
```

l. 

```{r echo = FALSE, fig.width = 4, fig.height = 2}
titanic <- read.csv("../data/titanic.csv", as.is = TRUE) %>%
        select(Survived, Pclass, Age, Name, Sex, Fare) %>% 
        mutate(Pclass = as.factor(Pclass))
titanic$agecat[titanic$Age < 15] <- "Under 15"          # 1. 
titanic$agecat[titanic$Age >= 15 & 
                     titanic$Age <= 50] <- "15--50"     # 2. 
titanic$agecat[titanic$Age > 50] <- "Over 50"
titanic$agecat <- factor(titanic$agecat,
                         levels = c("Under 15",
                                    "15--50",
                                    "Over 50"))

toplot_4 <- filter(titanic, !is.na(Age)) %>%            # 3. 
        select(Survived, Pclass, agecat, Sex) %>%       # 4. 
        group_by(Pclass, agecat, Sex) %>%
        summarize(N = n(),
                  survivors = sum(Survived == 1),
                  perc_survived = 100 * survivors / N) %>% #5. 
        ungroup() %>%
        mutate(Pclass = factor(Pclass, levels = c(1, 2, 3),
                               labels = c("First class", "Second class",
                                          "Third class"))) # 6.
plot_4 <- ggplot(toplot_4, aes(x = agecat, y = perc_survived, fill = Pclass)) + 
        geom_bar(stat = "identity", position = "dodge") +  # 7.
        coord_flip() +                                     # 8.
        ylab("Percent survivors") + xlab("Age category")
plot_4
```

> **Answers:**
> 1. b.,
> 2. g.,
> 3. d.,
> 4. e.,
> 5. c.,
> 6. k.,
> 7. h.,
> 8. j.,
> 9. l.,
> 10. f.,
> 11. i.,
> 12. a.