---
title: "Quiz 4"
author: ' '
date: ' '
output: pdf_document
---

```{r echo = FALSE, message = FALSE}
library(knitr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggthemes)
library(gridExtra)
knitr::opts_chunk$set(fig.path='figures/quiz4/plot-')
```

##### Name: 

#### 1.--4. 

You will be using the `worldcup` dataframe for some of the questions. As a reminder, here's what that looks like:

```{r}
library(faraway)
data(worldcup)
head(worldcup)
```

What would the head of the dataframe look like after each step of `dplyr` functions? Match the starts of the dataframes given in the numbered answers to the letter given in the comments following each line of the chained command:

```{r, results = "hide"}
my_df <- select(worldcup, Time, Passes, Tackles, Saves) %>%  # a.
        summarize(Time = mean(Time), 
                  Passes = mean(Passes), 
                  Tackles = mean(Tackles),
                  Saves = mean(Saves)) %>%  # b. 
        gather(var, mean) %>%  # c.
        mutate(mean = round(mean, 1))  # d. 
```

#### 1. 

```{r, echo = FALSE}
head(select(worldcup, Time, Passes, Tackles, Saves) %>%  # a.
        summarize(Time = mean(Time), 
                  Passes = mean(Passes), 
                  Tackles = mean(Tackles),
                  Saves = mean(Saves)) %>%  # b. 
        gather(var, mean))
```

#### 2. 

```{r echo = FALSE}
head(select(worldcup, Time, Passes, Tackles, Saves))
```

#### 3. 

```{r echo = FALSE}
head(select(worldcup, Time, Passes, Tackles, Saves) %>%  # a.
        summarize(Time = mean(Time), 
                  Passes = mean(Passes), 
                  Tackles = mean(Tackles),
                  Saves = mean(Saves)) %>%  # b. 
        gather(var, mean) %>%  # c.
        mutate(mean = round(mean, 1)))
```

#### 4. 

```{r, echo = FALSE}
head(select(worldcup, Time, Passes, Tackles, Saves) %>%  # a.
        summarize(Time = mean(Time), 
                  Passes = mean(Passes), 
                  Tackles = mean(Tackles),
                  Saves = mean(Saves)))
```

> **Answers:**
> 1. c., 
> 2. a.,
> 3. d.,
> 4. b. 

#### 5. 

You run the following code:

```{r, results = "hide"}
i <- c(1, 3:4)
j <- 2

worldcup[j, i]
```

What will you get?

a. `r worldcup[i, 1]`
b. `r worldcup[i, j]`
c. `r worldcup[j, i]`
d. `r worldcup[j, j]`

> **Answer: c**
> Explanation: This is the same as running `worldcup[2, c(1, 3:4)]`. R will look for the definitions of `i` and `j` and put this into the indexing. Therefore, this is calling for the second row, first, third, and fourth columns. 

#### 6. 

You run the following code:

```{r, results = "hide"}
vars <- c("Time", "Shots", "Passes", "Tackles", "Saves")
i <- 3

var_mean <- mean(worldcup[ , vars[i]])
var_mean
```

What will you get? (Hint: you may want to use some of the output shown in Questions 1--4 to help figure this out.)

a. `r mean(worldcup$Shots)`
b. `r mean(worldcup$Passes)`
c. `r mean(worldcup$Time)`
d. `r mean(worldcup$Saves)`

> **Answer: b**
> Explanation: Since `i` is defined as 3, `vars[i]` tells R to find the 3rd element of the `vars` vector, which is "Passes". `worldcup[ , vars[i]]` therefore pulls the "Passes" column of `worldcup`. The `mean` function takes the mean of that, which you can see from the output of questions 1--4 is `r mean(worldcup$Passes)`. 

#### 7.--10.

These questions plot data from another dataset in the `faraway` library, `hsb`. This includes data on the academic program choice of high school students. You can find out more using `?hsb`. 

```{r}
data(hsb)
head(hsb, 3)
```

Match up the graphs given by each number below with the letter following lines of code in the `ggplot` call. You should match each letter with the graph you would get if you ran *up to and including* the line of code the letter appears on, minus the `+` at the end of the line.

```{r, fig.keep='none'}
ggplot(hsb, aes(x = math, y = science)) + 
        geom_point(aes(shape = schtyp, color = schtyp)) + 
        scale_color_grey() +  # a.
        facet_wrap(~ prog, ncol = 1) +  # b. 
        xlab("Math score") + ylab("Science score") +  # c.
        theme_few()  # d.
```

```{r, echo = FALSE, fig.width = 10, fig.height = 8}
a <- ggplot(hsb, aes(x = math, y = science)) + 
        geom_point(aes(shape = schtyp, color = schtyp)) + 
        scale_color_grey() + 
        facet_wrap(~ prog, ncol = 1) + 
        xlab("Math score") + ylab("Science score") + 
        theme_few() + ggtitle("7.")

b <- ggplot(hsb, aes(x = math, y = science)) + 
        geom_point(aes(shape = schtyp, color = schtyp)) + 
        scale_color_grey() + 
        facet_wrap(~ prog, ncol = 1) + 
        xlab("Math score") + ylab("Science score") + ggtitle("8.")

c <- ggplot(hsb, aes(x = math, y = science)) + 
        geom_point(aes(shape = schtyp, color = schtyp)) + 
        scale_color_grey() + ggtitle("9.")

d <- ggplot(hsb, aes(x = math, y = science)) + 
        geom_point(aes(shape = schtyp, color = schtyp)) + 
        scale_color_grey() + 
        facet_wrap(~ prog, ncol = 1) + ggtitle("10.")

grid.arrange(a, b, c, d, ncol = 2)
```

> **Answers**:
> 7. d.,
> 8. c.,
> 9. a.,
> 10. b. 